A faster exact method for solving the robust multi-mode resource-constrained project scheduling problem

### 2022-03-11 paper/paper.tex
Complete first draft. Ready for Marc to review.

% bibliography and referencing
\usepackage{natbib}
\bibliographystyle{apalike}
\usepackage{bibentry}  % inline referencing
\renewcommand{\inlinebibentry}[1]{\cite{#1} \bibentry{#1}}
\usepackage{hyperref}
\usepackage{xcolor}
\definecolor{darkgreen}{RGB}{0,127,173}
\definecolor{darkblue}{RGB}{0,0,200}
\hypersetup{colorlinks, linkcolor={darkblue}, citecolor={darkgreen}, urlcolor={darkblue}}

@@

\nobibliography*  % tells bibentry to load bibliography database

@@

	The proposed mixed-integer linear programming formulation is an extension of the formulation proposed by \cite{bold2021compact} (\bibentry{bold2021compact}) for the problem without multiple activity modes. Computational experiments show that this easy-to-implement formulation can be solved many times faster than the current state-of-the-art for solving this problem, solving over 40\% more instances to optimality over the same set of benchmarking instances. 
\end{abstract}

@@

\section{Introduction}
The multi-mode resource-constrained project scheduling problem (MRCPSP) is a generalisation of the widely-studied resource-constrained project scheduling problem (RCPSP) to include multiple processing modes for each activity. The inclusion of these modes is to model situations in which there is more than one way of executing project activities, with each option having its own duration and resource requirements. The MRCPSP consists selecting the processing modes and start times for a given set of activities, subject to a set of precedence constraints and limited resource availability, with the objective of minimising the overall project duration, known as the makespan.

In this paper we consider the MRCPSP under uncertain activity durations and model it using a two-stage adjustable robust optimisation framework. In this setting, a first-stage problem is solved to determine activity mode selections and make activity sequencing decisions to resolve resource conflicts. Following this, the actual activity durations are realised and a complete schedule is computed. The aim of the two-stage robust MRCPSP is to find a feasible first-stage solution (i.e. mode selection and sequencing decisions) in order to minimise the realised worse-case makespan, as computed in the second-stage. We refer to this problem as the robust MRCPSP.

In recent years this two-stage robust optimisation approach has been applied to the RCPSP. First to use this approach were \cite{artigues2013robust}, who present an iterative scenario-relaxation algorithm for this problem with the objective of minimising the worst-case absolute regret. \cite{bruni2017adjustable} propose a Benders'-style decomposition approach for solving the robust RCPSP with the objective of minimising the worst-case project makespan. This work was extended in \cite{bruni2018computational}, which presents a computational study comparing an additional Benders' decomposition approach against a primal decomposition algorithm. Recently, \cite{bold2021compact} introduced a compact reformulation of the robust RCPSP and presented results which showed the superiority of that formulation over the iterative methods developed in those papers.

The application of the this two-stage robust optimisation approach for the MRCPSP is a more recently development. To the best of our knowledge, the only existing paper to consider this problem is \cite{balouka2021robust}, in which the Benders' decomposition approach used by \cite{bruni2017adjustable} for the robust RCPSP has been extended for application to the MRCPSP. Analogous to that work, the aim of this paper is to apply the compact reformulation first developed for the RCPSP by \cite{bold2021compact} to the MRCPSP. Specifically, we present a computational comparison of our compact reformulation approach and a strengthened version of the Benders' decomposition approach.

\section{Problem description}

A project consists of a set of non-preemptive activities $V=\{0,1,\dots,n,n+1\}$, where 0 and $n+1$ denote the dummy source and sink activities respectively. Each activity $i\in V$ has a set of available processing modes given by $M_i=\{1,\dots,|M_i|\}$. The nominal duration of activity $i$ when executed in mode $m_i\in M_i$ is given by $\bar{d}_{im_i}$, whilst its worst-case duration is given by $\bar{d}_{im_i}+\hat{d}_{im_i}$, where $\hat{d}_{im_i}$ is its maximum durational deviation. Each mode for activity $i$, $m_i\in M_i$, has an associated renewable resource requirement of $r_{im_ik}$ for each $k\in K$, where $K$ is the set of renewable resource types involved in the project. Each renewable resource $k\in K$ has an availability of $R_{k}$ at each time period in the project horizon. As well as renewable resource requirements, mode $m_i$ of activity $i$ also has a non-renewable resource requirement of $r^{'}_{im_ik}$ for each non-renewable resource type $k\in K^{'}$, with each non-renewable resource having an overall availability of $R^{'}_k$ for the entire project horizon. Additionally, the project is subject to a set of strict finish-to-start precedence constraints given by $E$, where $(i,j)\in E$ enforces that activity $i$ must finish before activity $j$ can begin. These form a project network that can be represented using a directed graph $G(V,E)$. Figure \ref{fig:mrcpsp_network} shows an example instance involving five non-dummy activities and a single renewable resource.

\begin{figure}[h]
	\centering
	\includegraphics[scale=1]{mrcpsp_network.pdf}
	\caption{An example instance involving five non-dummy activities and a single renewable resource. Activities 2 and 5 each have two available processing modes, whilst the other activities have only a single mode.}
	\label{fig:mrcpsp_network}
\end{figure}

For a given choice of processing modes for each activity $\bm{m}=(m_1,\dots,m_n)$, we assume that the activity durations lie somewhere in a budgeted uncertainty set of the form
$$\U_{\bm{m}}(\Gamma)=\Bigg\{\bm{d}\in\mathbb{R}_+^{|V|}:d_{im_i}=\bar{d}_{im_i}+\xi_i\hat{d}_{im_i},\,0\leq \xi_i \leq 1,\,\forall i\in V,\,\sum_{i\in V}\xi_i \leq \Gamma\Bigg\}.$$
Introduced by \cite{bertsimas2004price}, the motivation of the budgeted uncertainty set is to control the pessimism of the solution by introducing a robustness parameter $\Gamma$ to limit the number of jobs that can simultaneously achieve their worst-case durations. Observe that when $\Gamma=0$, each activity takes its nominal duration and the resulting problem is the deterministic MRCPSP. At the other extreme, when $\Gamma=n$, every activity takes its worst-case duration, in which case the uncertainty set becomes equivalent to an interval uncertainty set. In this case the problem can again be solved as a deterministic MRCPSP instance considering only worst-case durations.

For a given set of activity modes $\bm{m}$, a forbidden set is defined to be any subset of activities $F_{\bm{m}}\subseteq V$ that are not precedence-related, such that $\sum_{i\in F_{\bm{m}}}r_{im_ik}>R_k$ for at least one resource $k\in K$. That is, a forbidden set is a collection of activities that cannot be executed in parallel only because of resource limitations.

Applying the main representation theorem of \cite{bartusch1988scheduling}, for a particular choice of activity modes $\bm{m}$, a solution to the MRCPSP can be defined by a set of additional precedences $X_{\bm{m}}\subseteq V^2\setminus E$ such that the extended precedence network $G(V,E\cup X_{\bm{m}})$ is acyclic and contains no forbidden sets. Such an extension to the project network is referred to as a \textit{sufficient selection}.

Hence the two-stage robust MRCPSP under budgeted uncertainty, which we refer to as the robust MRCPSP, can be written as
\begin{align}
	& \min_{\bm{m}\in \M,\,X_{\bm{m}}\in \X_{\bm{m}}} \max_{\bm{d}\in\U_{\bm{m}}(\Gamma)} \min S_{n+1}\label{eqn:robust_mrcpsp_1}\\
	& S_0 = 0\\
	& S_j - S_i \geq d_{im_i} \qquad \forall (i,j)\in E\cup X_{\bm{m}}\\
	& S_i \geq 0 \qquad \forall i\in V,\label{eqn:robust_mrcpsp_4}
\end{align}
where $\M\subseteq \mathbb{N}^n$ represents the set of all possible combinations of activity processing mode selections, and $\X_{\bm{m}}$ is the set of all possible sufficient selections for the choice of processing modes given by $\bm{m}$. This problem aims to selects activity modes and a corresponding sufficient selection in order to minimise the worst-case project makespan.

\section{A compact formulation}

We propose solving the robust MRCPSP using an extended version of the mixed-integer programming formulation developed by \cite{bold2021compact} for solving the robust RCPSP. This formulation combines the first and second-stage problems into a single compact formulation and was shown to significantly outperform the strongest iterative decomposition-based methods for the robust RCPSP.

This formulation is constructed by recasting the adversarial subproblem of determining the worst-case activity durations for a given set of mode choices and sufficient selection as a longest path problem on an augmented project network. This augmented project network is formed of $\Gamma+1$ connected copies of the original project network, with each level being used to account for a delay to a single activity in the project. Having reformulated the adversarial subproblem as a longest-path problem, it can be dualised an inserted into a formulation for the first-stage problem, to arrive at a complete compact formulation for the full problem. Given that the derivation of this formulation is more or less identical to the derivation of the corresponding formulation for the robust RCPSP, we omit it here and instead refer the reader to \cite{bold2021compact}.

In the resulting compact formulation, $y_{ij}$ variables are used to define a complete set of transitive precedences between the project activities that are formed by the original project precedence constraints and the additional precedences introduced to resolve resource conflicts, i.e. $\{(i,j),\, i,j\in V:y_{ij}=1\}=T(E\cup X_{\bm{m}})$, where $T(\cdot)$ is used to denote the transitive closure of a set. Variables $x_{im}$ determine the processing mode for each activity, whilst resource flow variables $f_{ijk}$ track the amount of renewable resource $k$ that is transferred from activity $i$ to activity $j$ upon its completion. The $S_{i\gamma}$ variables are dual variables associated with the longest-path subproblem, which in this formulation these variables are used to track the start time of activities under the worst-case activity durations. Using these variables, the compact formulation for the robust MRCPSP can be written as   
\begin{align}
\min \ & S_{n+1,\Gamma}\label{eqn:compact1}\\
\text{s.t. } & S_{00}=0\label{eqn:compact2}\\
	     & S_{j\gamma} - S_{i\gamma} \geq \bar{d}_{im}x_{im} - N(1-y_{ij}) \qquad \forall (i,j)\in V^2,\,\forall m\in M_i,\,\gamma=0,\dots,\Gamma\label{eqn:compact_bigM1}\\
	     & S_{j,\gamma+1} - S_{i\gamma} \geq (\bar{d}_{im} + \hat{d}_{im})x_{im} - N(1-y_{ij})\nonumber \\
	     & & \hspace{-8cm} \forall (i,j)\in V^2,\, m\in M_i,\,\gamma = 0,\dots,\Gamma-1\label{eqn:compact_bigM2}\\
	     & y_{ij}=1 \qquad \forall (i,j) \in E\cup\{(n+1,n+1)\}\label{eqn:compact5}\\
	     & y_{ij} + y_{ji} \leq 1 \qquad \forall i,j\in V,\, i<j\label{eqn:compact_trans1}\\
	     & y_{ij} = y_{ip} + y_{p j} - 1 \qquad \forall i,j,p\in V,\,i\neq j\neq p\label{eqn:compact_trans2}\\
	     & f_{ijk}\leq P_{ijk}y_{ij} \qquad \forall (i,j)\in V^2,\,i\neq n+1,\,j\neq0,\, \forall k \in K\label{eqn:compact6}\\
	     & \sum_{i\in V\setminus\{n+1\}}f_{ijk}=\sum_{m\in M_j}r_{jmk}x_{jm} \qquad \forall j \in V\setminus\{0\} ,\, \forall k\in K\label{eqn:compact7}\\
	     & \sum_{j\in V\setminus\{0\}}f_{ijk}=\sum_{m\in M_i}r_{imk}x_{im} \qquad \forall i \in V\setminus\{n+1\},\,\forall k\in K\label{eqn:compact8}\\
	     & \sum_{m\in M_i}x_{im} = 1 \qquad \forall i\in V\label{eqn:compact9}\\
	     & \sum_{m\in M_i}r^{'}_{imk}x_{im} \leq R^{'}_k \qquad \forall i \in V,\,k\in K^{'}\label{eqn:compact_nonrenewable}\\
	     & S_{i\gamma}\geq 0 \qquad \forall i\in V,\,\gamma \in 0,\dots, \Gamma\label{eqn:compact10}\\
	     & y_{ij}\in\{0,1\} \qquad \forall (i,j)\in V^2\label{eqn:compact11}\\
	     & f_{ijk}\geq 0 \qquad \forall (i,j)\in V^2 ,\, \forall k \in K\label{eqn:compact12}\\
	     & x_{im}\in\{0,1\} \qquad \forall i\in V,\,m\in M_i\label{eqn:compact13},
\end{align}
where $N=\sum_{i\in V}\max_{m\in M_i}(\bar{d}_{im}+\hat{d}_{im})$ is an upper bound on the minimum makespan, and $P_{ijk}=\min\{\max_{m\in M_i}r_{imk},\,\max_{m\in M_j} r_{jmk}\}$ is the maximum possible flow of resource $k$ from $i$ into $j$.

@@

\section{A Benders' decomposition approach}

In this section we outline an alternative approach to solving the robust MRCPSP based on Benders' decomposition. Such a solution method for the robust MRCPSP was first presented in \cite{balouka2021robust}, and as mentioned previously, this itself was based on the approach to solve the robust RCPSP used in \cite{bruni2017adjustable}.

The main idea behind this approach is to decompose the full problem into a two smaller problems: 1. a master problem that determines activity processing modes and resolves resource conflicts, and 2. a subproblem that takes the solution from the master problem and evaluates it by finding the worst-case makespan for the resulting network by solving a longest path problem. The solution to the master problem forms a lower bound to the optimal objective value of the original problem, whilst the solution to the subproblem provides an upper bound. These two problems are solved iteratively, with optimality cuts being added to the master problem each iteration based on the solution to the subproblem until the lower and upper bounds meet.

After initially implementing the Benders' algorithm as it is presented in \cite{balouka2021robust}, it was discovered that its performance could be strengthened by replacing the master problem formulation used in that implementation with a stripped-down version of the compact formulation (\ref{eqn:compact1})-(\ref{eqn:compact13}) in which the uncertainty is removed. In this section we therefore present our strengthened implementation of the Benders' approach for solving the robust MRCPSP. Note that a comparison of these two Benders' implementations is included in Section \ref{section:experiments}. For the details of the original implementation of the Benders' decomposition algorithm for the robust MRCPSP, see \cite{balouka2021robust}.

@@

The subproblem at iteration $t$ considers the worst-case makespan for the mode choices and sufficient selection from the master problem at iteration $t$, denoted by $\bm{m}^{*t}$ and $X_{\bm{m}^{*t}}$ respectively. This is done by solving an uncertain longest-path problem through the extended project network $T(E\cup X_{\bm{m}^{*t}})$, where activity durations lie in the budgeted uncertainty set $\U_{\bm{m}^{*t}}(\Gamma)$, which can be formulated as 
\begin{align}
	V^{*t} = &\max \sum_{(i,j)\in T(E\cup X_{\bm{m}^{*t}})} \bar{d}_{im^{*t}_i}\alpha_{ij} + \hat{d}_{im^{*t}_i}w_{ij}\label{eqn:subproblem_first}\\
	& \sum_{(i,n+1)\in T(E\cup X_{\bm{m}^{*t}})} \alpha_{i,n+1} = 1\\
	& \sum_{(0,i)\in T(E\cup X_{\bm{m}^{*t}})} \alpha_{0,i} = 1\\
	& \sum_{(i,j)\in T(E\cup X_{\bm{m}^{*t}})} \alpha_{ij} - \sum_{(j,i)\in T(E\cup X_{\bm{m}^{*t}})} \alpha_{ji} = 0 \qquad \forall i \in V\setminus\{0,n+1\}\\
	& w_{ij} \leq \xi_i \qquad \forall (i,j)\in T(E\cup X_{\bm{m}^{*t}})\\
	& w_{ij} \leq \alpha_{ij} \qquad \forall (i,j)\in T(E\cup X_{\bm{m}^{*t}})\\
	& \sum_{i\in V}\xi_i \leq \Gamma\\
	& \alpha_{ij}\in \{0,1\} \qquad \forall (i,j)\in T(E\cup X_{\bm{m}^{*t}})\\
	& w_{ij}\geq 0 \qquad \forall (i,j)\in T(E\cup X_{\bm{m}^{*t}})\\
	& 0\leq \xi_i \leq 1 \qquad \forall i\in V. \label{eqn:subproblem_last}
\end{align}

Variables $\alpha_{ij}$ define the longest path through the network. This path, which is denote as $\pi^{*t}=\{(i,j),\,i,j\in V:\alpha_{ij}=1\}$ has length $V^{*t}$. Variables $\xi_i$ are the activity delay variables, whilst $w_{ij}$ are used to linearise the formulation. Note that this formulation of the subproblem is identical to the one implemented by \cite{balouka2021robust}.

@@

Given the current best lower bound for the original problem $LB$, the mode selection from the master problem $\bm{m}^{*t}$, and the optimal objective value of the subproblem $V^{*t}$, the cut at iteration $t$ can be written as
\begin{align}
	& \eta \geq (V^{*t} - LB)\cdot\sum_{(i,j)\in \pi^{*t}}\bigg(1/3 (y_{ij} + x_{i,m_i^{*t}} + x_{j,m_j^{*t}}) - (3 - y_{ij} - x_{i,m_i^{*t}} - x_{j,m_j^{*t}})\bigg) & \nonumber \\
	& \hspace{8.1cm} - (V^{*t} - LB)\cdot(|\pi^{*t}|-1) + LB,\label{eqn:benders_cut}
\end{align}

@@

>>>>

% bibliography and referencing
\usepackage{natbib}
\bibliographystyle{apalike}
\usepackage{hyperref}
\usepackage{xcolor}
\definecolor{darkgreen}{RGB}{0,127,173}
\definecolor{darkblue}{RGB}{0,0,200}
\hypersetup{colorlinks, linkcolor={darkblue}, citecolor={darkgreen}, urlcolor={darkblue}}

@@

@@

Computational experiments show that this easy-to-implement formulation is many times faster to solve than the current Benders' decomposition-based state-of-the-art solution approach for this problem, whilst also being able to solve over 40\% more instances to optimality over the same set of benchmarking instances.
\end{abstract}

@@

\section{Introduction}
The multi-mode resource-constrained project scheduling problem (MRCPSP) is a generalisation of the widely-studied resource-constrained project scheduling problem (RCPSP) to include multiple processing modes for each activity. The inclusion of these modes allows the modelling of situations in which there is more than one way of executing project activities, with each option having its own duration and resource requirements. The MRCPSP consists of selecting the processing modes and start times for a given set of activities, subject to a set of precedence constraints and limited resource availability, with the objective of minimising the overall project duration, known as the makespan.

In this paper we consider the MRCPSP under uncertain activity durations and model it using a two-stage adjustable robust optimisation framework. In this setting, a first-stage problem is solved to determine activity mode selections and make activity sequencing decisions to resolve resource conflicts. Following this, the actual activity durations are realised and a complete schedule is computed. The aim of the two-stage adjustable robust MRCPSP is to find a feasible first-stage solution (i.e. mode selection and sequencing decisions) in order to minimise the realised worse-case makespan, as computed in the second-stage. We refer to this problem as the robust MRCPSP.

A number of papers in recent years have applied this two-stage robust optimisation approach to the RCPSP. First to use this approach were \cite{artigues2013robust}, who presented an iterative scenario-relaxation algorithm for this problem with the objective of minimising the worst-case absolute regret. More recently, \cite{bruni2017adjustable} introduced a Benders'-style decomposition approach for solving the robust RCPSP with the objective of minimising the worst-case project makespan. This work was extended in \cite{bruni2018computational}, which presented a computational study comparing an additional Benders' decomposition approach against a primal decomposition algorithm. Most recently, \cite{bold2021compact} introduced a compact reformulation of the robust RCPSP and presented results which showed the superior computational performance of that formulation over the iterative decomposition-based methods developed in the two preceding papers.

The application of this two-stage robust optimisation approach for the MRCPSP is a very recent development. To the best of our knowledge, the only existing paper to consider this problem is \cite{balouka2021robust}, in which the Benders' decomposition approach introduced by \cite{bruni2017adjustable} for the robust RCPSP has been extended for application to the MRCPSP. Mirroring that extension, this paper adapts the compact formulation developed by \cite{bold2021compact} to the MRCPSP, with the aim of achieving similarly superior computational performance over the Benders' solution approach.

Following a formal description of the two-stage robust MRCPSP in Section \ref{section:problem_description}, we outline the proposed compact formulation for this problem in Section \ref{section:compact_formulation} and present a strengthened version of the Benders' decomposition solution approach from \cite{balouka2021robust} in Section \ref{section:benders}. Results from a computational comparison of these two approaches are detailed in Section \ref{section:experiments}.

\section{Problem description}\label{section:problem_description}

A project consists of a set of non-preemptive activities $V=\{0,1,\dots,n,n+1\}$, where 0 and $n+1$ denote the dummy source and sink activities respectively. Each activity $i\in V$ has a set of available processing modes given by $M_i=\{1,\dots,|M_i|\}$. The nominal duration of activity $i$ when executed in mode $m\in M_i$ is given by $\bar{d}_{im}$, whilst its worst-case duration is given by $\bar{d}_{im}+\hat{d}_{im}$, where $\hat{d}_{im}$ is its maximum durational deviation. Each mode for activity $i$, $m\in M_i$, has an associated renewable resource requirement of $r_{imk}$ for each $k\in K$, where $K$ is the set of renewable resource types involved in the project. Each renewable resource $k\in K$ has an availability of $R_{k}$ at each time period in the project horizon. As well as renewable resource requirements, mode $m$ of activity $i$ also has a non-renewable resource requirement of $r^{'}_{imk}$ for each non-renewable resource type $k\in K^{'}$, with each non-renewable resource having an overall availability of $R^{'}_k$ for the entire project horizon. Additionally, the project is subject to a set of strict finish-to-start precedence constraints given by $E$, where $(i,j)\in E$ enforces that activity $i$ must finish before activity $j$ can begin. These form a project network that can be represented using a directed graph $G(V,E)$. Figure \ref{fig:mrcpsp_network} shows an example instance involving five non-dummy activities and a single renewable resource.

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.9]{mrcpsp_network.pdf}
	\caption{An example instance involving five non-dummy activities and a single renewable resource. Activities 2 and 5 each have two available processing modes, whilst the other activities have only a single mode.}
	\label{fig:mrcpsp_network}
\end{figure}

For a given choice of processing modes for each activity $\bm{m}=(m_1,\dots,m_n)$, we assume that the activity durations lie somewhere in a budgeted uncertainty set of the form
$$\U_{\bm{m}}(\Gamma)=\Bigg\{\bm{d}_{\bm{m}}\in\mathbb{R}_+^{|V|}:d_{im_i}=\bar{d}_{im_i}+\xi_i\hat{d}_{im_i},\,0\leq \xi_i \leq 1,\,\forall i\in V,\,\sum_{i\in V}\xi_i \leq \Gamma\Bigg\}.$$
Introduced by \cite{bertsimas2004price}, the motivation of the budgeted uncertainty set is to control the pessimism of the solution by introducing a robustness parameter $\Gamma$ to limit the number of jobs that can simultaneously achieve their worst-case durations. Observe that when $\Gamma=0$, each activity takes its nominal duration and the resulting problem is the deterministic MRCPSP. At the other extreme, when $\Gamma=n$, every activity takes its worst-case duration, in which case the uncertainty set becomes equivalent to an interval uncertainty set. In this case the problem can again be solved as a deterministic MRCPSP instance considering only worst-case durations.

For a given set of activity modes $\bm{m}$, a forbidden set is defined to be any subset of activities $F_{\bm{m}}\subseteq V$ that are not precedence-related, such that $\sum_{i\in F_{\bm{m}}}r_{im_ik}>R_k$ for at least one resource $k\in K$. That is, a forbidden set is a collection of activities that cannot be executed in parallel only because of resource limitations. Applying the main representation theorem of \cite{bartusch1988scheduling}, for a particular choice of activity modes $\bm{m}$, a solution to the MRCPSP can be defined by a set of additional precedences $X_{\bm{m}}\subseteq V^2\setminus E$ such that the extended precedence network $G(V,E\cup X_{\bm{m}})$ is acyclic and contains no forbidden sets. Such an extension to the project network is referred to as a \textit{sufficient selection}.

The aim of the two-stage robust MRCPSP is therefore to select set of activity modes and a corresponding sufficient selection in order to minimise the worst-case project makespan. For the case where activity durations lie in a budgeted uncertainty set that we consider in this paper, this problem can be written as
\begin{align}
	& \min_{\bm{m}\in \M,\,X_{\bm{m}}\in \X_{\bm{m}}} \max_{\bm{d}\in\U_{\bm{m}}(\Gamma)} \min S_{n+1}\label{eqn:robust_mrcpsp_1}\\
	& S_0 = 0\\
	& S_j - S_i \geq d_{im_i} \qquad \forall (i,j)\in E\cup X_{\bm{m}}\\
	& S_i \geq 0 \qquad \forall i\in V,\label{eqn:robust_mrcpsp_4}
\end{align}
where $\M\subseteq \mathbb{N}^n$ represents the set of all possible combinations of activity processing mode selections, and $\X_{\bm{m}}$ is the set of all possible sufficient selections for the choice of processing modes given by $\bm{m}$.

An optimal solution to the instance shown in Figure \ref{fig:mrcpsp_network} is given by Figure \ref{fig:mrcpsp_network_solution}, where the mode choices are shown in bold, and the sufficient selection is given by $\{(3,2)\}$. The worst-case schedule for this optimal solution is shown in Figure \ref{fig:mrcpsp_schedule}, where activities 1 and 3 have been delayed to achieve a worst-case makespan of 15 (note that delaying a combination of activity 1 and any other activity results in the same worst-case makespan).

\begin{figure}[htb]
	\makebox[\textwidth][c]{
	\begin{subfigure}[t]{.5\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{mrcpsp_solution.pdf}
		\caption{Activity mode choices are highlighted in bold and sufficient selection is given by dashed arc $\{(3,2)\}$.}\label{fig:mrcpsp_network_solution}
	\end{subfigure}%
	\hspace{15mm}
	\begin{subfigure}[t]{.6\textwidth}
		\centering
		\includegraphics[width=1.00\textwidth]{mrcpsp_schedule.pdf}
		\caption{Worst-case schedule corresponding to optimal robust solution, where activities 1 and 3 have been delayed.}\label{fig:mrcpsp_schedule}
	\end{subfigure}
	}
	\caption{Optimal solution to the example instance from Figure \ref{fig:mrcpsp_network}.}
	\label{fig:mrcpsp_solution}
\end{figure}

\section{A compact formulation}\label{section:compact_formulation}

We propose solving the robust MRCPSP using an extended version of the mixed-integer programming formulation developed by \cite{bold2021compact} for solving the two-stage adjustable robust RCPSP. This formulation combines the first and second-stage problems into a single compact formulation and was shown to significantly outperform the strongest iterative decomposition-based methods for that problem.

This formulation is constructed by recasting the adversarial subproblem of determining the worst-case activity durations for a given set of mode choices and sufficient selection as a longest path problem on an augmented project network. This augmented project network is formed of $\Gamma+1$ connected copies of the original project network, with each level being used to account for a delay to a single activity in the project. Having reformulated the adversarial subproblem as a longest-path problem, strong duality can be employed to enable its insertion it into a formulation for the first-stage problem, and ultimately arrive at a complete compact formulation for the full problem. Given that the derivation of this formulation is more or less identical to the derivation of the corresponding formulation for the robust RCPSP, we omit it here and instead refer the reader to \cite{bold2021compact}.

In the resulting compact formulation, $y_{ij}$ variables are used to define a complete set of transitive precedences between the project activities that are formed by the original project precedence constraints and the additional precedences introduced to resolve resource conflicts, i.e. $\{(i,j),\, i,j\in V:y_{ij}=1\}=T(E\cup X_{\bm{m}})$, where $T(\cdot)$ is used to denote the transitive closure of a set. Variables $x_{im}$ determine the processing mode for each activity, whilst resource flow variables $f_{ijk}$ track the amount of renewable resource $k$ that is transferred from activity $i$ to activity $j$ upon its completion. The $S_{i\gamma}$ variables are dual variables associated with the longest-path subproblem, which in this formulation are used to track the start time of activities under the worst-case activity durations. Using these variables, the compact formulation for the robust MRCPSP can be written as   
\begin{align}
\min \ & S_{n+1,\Gamma}\label{eqn:compact1}\\
\text{s.t. } & S_{00}=0\label{eqn:compact2}\\
	     & S_{j\gamma} - S_{i\gamma} \geq \bar{d}_{im}x_{im} - N(1-y_{ij}) \qquad \forall (i,j)\in V^2,\,\forall m\in M_i,\,\gamma=0,\dots,\Gamma\label{eqn:compact_bigM1}\\
	     & S_{j,\gamma+1} - S_{i\gamma} \geq (\bar{d}_{im} + \hat{d}_{im})x_{im} - N(1-y_{ij})\nonumber \\
	     & & \hspace{-8cm} \forall (i,j)\in V^2,\, m\in M_i,\,\gamma = 0,\dots,\Gamma-1\label{eqn:compact_bigM2}\\
	     & y_{ij}=1 \qquad \forall (i,j) \in E\cup\{(n+1,n+1)\}\label{eqn:compact5}\\
	     & y_{ij} + y_{ji} \leq 1 \qquad \forall i,j\in V,\, i<j\label{eqn:compact_trans1}\\
	     & y_{ij} = y_{ip} + y_{p j} - 1 \qquad \forall i,j,p\in V,\,i\neq j\neq p\label{eqn:compact_trans2}\\
	     & f_{ijk}\leq P_{ijk}y_{ij} \qquad \forall (i,j)\in V^2,\,i\neq n+1,\,j\neq0,\, \forall k \in K\label{eqn:compact6}\\
	     & \sum_{i\in V\setminus\{n+1\}}f_{ijk}=\sum_{m\in M_j}r_{jmk}x_{jm} \qquad \forall j \in V\setminus\{0\} ,\, \forall k\in K\label{eqn:compact7}\\
	     & \sum_{j\in V\setminus\{0\}}f_{ijk}=\sum_{m\in M_i}r_{imk}x_{im} \qquad \forall i \in V\setminus\{n+1\},\,\forall k\in K\label{eqn:compact8}\\
	     & \sum_{m\in M_i}x_{im} = 1 \qquad \forall i\in V\label{eqn:compact9}\\
	     & \sum_{m\in M_i}r^{'}_{imk}x_{im} \leq R^{'}_k \qquad \forall i \in V,\,k\in K^{'}\label{eqn:compact_nonrenewable}\\
	     & S_{i\gamma}\geq 0 \qquad \forall i\in V,\,\gamma \in 0,\dots, \Gamma\label{eqn:compact10}\\
	     & y_{ij}\in\{0,1\} \qquad \forall (i,j)\in V^2\label{eqn:compact11}\\
	     & f_{ijk}\geq 0 \qquad \forall (i,j)\in V^2 ,\, \forall k \in K\label{eqn:compact12}\\
	     & x_{im}\in\{0,1\} \qquad \forall i\in V,\,m\in M_i\label{eqn:compact13},
\end{align}
where $N=\sum_{i\in V}\max_{m\in M_i}(\bar{d}_{im}+\hat{d}_{im})$ is an upper bound on the minimum makespan, and $P_{ijk}=\min\{\max_{m\in M_i}r_{imk},\,\max_{m\in M_j} r_{jmk}\}$ is the maximum possible flow of resource $k$ from $i$ into $j$.

@@

\section{A Benders' decomposition approach}\label{section:benders}

In this section we outline an alternative approach to solving the robust MRCPSP based on Benders' decomposition. A Benders'-type approach for solving the robust MRCPSP was first presented in \cite{balouka2021robust}, and as mentioned previously, this itself was based on the approach for solving the robust RCPSP used in \cite{bruni2017adjustable}.

The main idea behind this approach is to decompose the full problem into its two stages: 1. a master problem that determines activity processing modes and resolves resource conflicts, and 2. a subproblem that takes the solution from the master problem and evaluates it by finding the worst-case makespan for the resulting network by solving a longest path problem. Since the master problem does not account for all the problem uncertainty at once, its solution forms a lower bound to the optimal objective value of the original problem. Meanwhile, the solution to the subproblem is feasible to the original problem, and therefore provides an upper bound. These two problems are solved iteratively, with optimality cuts being added to the master problem each iteration based on the solution to the subproblem until the lower and upper bounds meet.

After initially implementing the Benders' algorithm as it is presented in \cite{balouka2021robust}, it was discovered that its performance could be strengthened by replacing the master problem formulation used in that implementation with a stripped-down version of the compact formulation (\ref{eqn:compact1})-(\ref{eqn:compact13}) in which the uncertainty is removed. In this section we present our strengthened implementation of the Benders' approach for solving the robust MRCPSP. A comparison of these two Benders' implementations is included in the results in Section \ref{section:experiments}. For the details of the original implementation of the Benders' decomposition algorithm for the robust MRCPSP, see \cite{balouka2021robust}.

@@

The subproblem at iteration $t$ computes the worst-case makespan for the mode choices and sufficient selection from the master problem at iteration $t$, denoted by $\bm{m}^{*t}$ and $X_{\bm{m}^{*t}}$ respectively. This is done by solving a longest-path problem through the extended project network $T(E\cup X_{\bm{m}^{*t}})$, in which the activity durations (i.e. arc lengths) can be chosen from the budgeted uncertainty set $\U_{\bm{m}^{*t}}(\Gamma)$. This problem can be formulated as 
\begin{align}
	V^{*t} = &\max \sum_{(i,j)\in T(E\cup X_{\bm{m}^{*t}})} \bar{d}_{im^{*t}_i}\alpha_{ij} + \hat{d}_{im^{*t}_i}w_{ij}\label{eqn:subproblem_first}\\
	& \sum_{(i,n+1)\in T(E\cup X_{\bm{m}^{*t}})} \alpha_{i,n+1} = 1\\
	& \sum_{(0,i)\in T(E\cup X_{\bm{m}^{*t}})} \alpha_{0,i} = 1\\
	& \sum_{(i,j)\in T(E\cup X_{\bm{m}^{*t}})} \alpha_{ij} - \sum_{(j,i)\in T(E\cup X_{\bm{m}^{*t}})} \alpha_{ji} = 0 \qquad \forall i \in V\setminus\{0,n+1\}\\
	& w_{ij} \leq \xi_i \qquad \forall (i,j)\in T(E\cup X_{\bm{m}^{*t}})\\
	& w_{ij} \leq \alpha_{ij} \qquad \forall (i,j)\in T(E\cup X_{\bm{m}^{*t}})\\
	& \sum_{i\in V}\xi_i \leq \Gamma\\
	& \alpha_{ij}\in \{0,1\} \qquad \forall (i,j)\in T(E\cup X_{\bm{m}^{*t}})\\
	& w_{ij}\geq 0 \qquad \forall (i,j)\in T(E\cup X_{\bm{m}^{*t}})\\
	& 0\leq \xi_i \leq 1 \qquad \forall i\in V. \label{eqn:subproblem_last}
\end{align}

Variables $\alpha_{ij}$ define the longest path through the network, which we denote as $\pi^{*t}=\{(i,j),\,i,j\in V:\alpha_{ij}=1\}$, and has length $V^{*t}$. Variables $\xi_i$ are the activity delay variables, whilst $w_{ij}$ are used to linearise the formulation. Note that this formulation of the subproblem is identical to the one implemented by \cite{balouka2021robust}.

@@

Given the current best lower bound for the original problem, $LB$, the mode selection from the master problem, $\bm{m}^{*t}$, and the optimal objective value of the subproblem, $V^{*t}$, the cut at iteration $t$ can be written as
\begin{align}
	& \eta \geq (V^{*t} - LB)\cdot\sum_{(i,j)\in \pi^{*t}}\bigg(1/3 (y_{ij} + x_{i,m_i^{*t}} + x_{j,m_j^{*t}}) - (3 - y_{ij} - x_{i,m_i^{*t}} - x_{j,m_j^{*t}})\bigg) & \nonumber \\
	& \hspace{8.1cm} - (V^{*t} - LB)\cdot(|\pi^{*t}|-1) + LB.\label{eqn:benders_cut}
\end{align}

@@

An overview of the implementation of the Benders' solution approach outlined in this section is presented in Algorithm \ref{alg:benders}.
